Research Notes

### 2018-07-24 writeups/notes.tex
use analytical estimate for F:Uz1 to plot, init nonlin sims

    \newcommand*{\scinot}[2]{#1\times10^{#2}}
    \newcommand*{\bra}[1]{\left<#1\right|}
    \newcommand*{\ket}[1]{\left|#1\right>}
    \newcommand*{\dotp}[2]{\left<#1\,\middle|\,#2\right>}
    \newcommand*{\rd}[2]{\frac{\mathrm{d}#1}{\mathrm{d}#2}}
    \newcommand*{\pd}[2]{\frac{\partial#1}{\partial#2}}
    \newcommand*{\rtd}[2]{\frac{\mathrm{d}^2#1}{\mathrm{d}#2^2}}
    \newcommand*{\ptd}[2]{\frac{\partial^2 #1}{\partial#2^2}}
    \newcommand*{\md}[2]{\frac{\mathrm{D}#1}{\mathrm{D}#2}}
    \newcommand*{\norm}[1]{\left|\left|#1\right|\right|}
    \newcommand*{\abs}[1]{\left|#1\right|}
    \newcommand*{\pvec}[1]{\vec{#1}^{\,\prime}}
    \newcommand*{\svec}[1]{\vec{#1}\;\!}
    \newcommand*{\bm}[1]{\boldsymbol{\mathbf{#1}}}
    \newcommand*{\expvalue}[1]{\left<#1\right>}
    \newcommand*{\ang}[0]{\text{\AA}}
    \newcommand*{\mum}[0]{\upmu \mathrm{m}}
    \newcommand*{\at}[1]{\left.#1\right|}

@@

    \DeclarePairedDelimiter\p{\lparen}{\rparen}
    \DeclarePairedDelimiter\s{\lbrack}{\rbrack}
    \DeclarePairedDelimiter\z{\lbrace}{\rbrace}

@@

Another formula given by Sutherland 2011 is $\pd{\expvalue{E}}{t} = v_{g, z}
\expvalue{E}$ where we average over $x$ one wavelength. I wasn't able to show
that these two agree in general; Sutherland's formula seems to be derived for a
traveling wavepacket whereas we have no such thing. We do not use this
expression since the Landau \& Lifshitz expression works well.

@@

One can imagine, for this matter, that $\rd{}{t}$ can be wrapped in the $L_{ij}$
operator and the timestepping computed implicitly; I believe this is what
Dedalus does, implicitly timestepping the left hand side and explicitly the
inhomogeneous terms.

Both of these procedures require $\bm{L}$ linear operator or $\rd{}{t} + \bm{L}$
to be full rank, or invertible. This is enforced analytically by requiring the
same number of BCs as derivatives, and numerically this is little different.
This is why it is possible to have components of $Q$ that are not differentiated
$\partial_t$ but still have well-defined time evolutions; the only condition
required is that the linear terms of the PDE make up a full-rank operator.
Hence, in our linear incompressible equations, we permitted use of $\vec{\nabla}
\cdot \vec{u}_1 = 0$ in lieu of a $\pd{P}{t}$ explicit equation of motion; so
long as the gauge for $P$ (which only appears in derivatives) is properly set,
the operator is full rank and numerically well-defined.

@@

@@

@@

>>>>

    \newcommand*{\scinot}[2]{#1\times10^{#2}}
    \newcommand*{\dotp}[2]{\left<#1\,\middle|\,#2\right>}
    \newcommand*{\rd}[2]{\frac{\mathrm{d}#1}{\mathrm{d}#2}}
    \newcommand*{\pd}[2]{\frac{\partial#1}{\partial#2}}
    \newcommand*{\rtd}[2]{\frac{\mathrm{d}^2#1}{\mathrm{d}#2^2}}
    \newcommand*{\ptd}[2]{\frac{\partial^2 #1}{\partial#2^2}}
    \newcommand*{\md}[2]{\frac{\mathrm{D}#1}{\mathrm{D}#2}}
    \newcommand*{\pvec}[1]{\vec{#1}^{\,\prime}}
    \newcommand*{\svec}[1]{\vec{#1}\;\!}
    \newcommand*{\bm}[1]{\boldsymbol{\mathbf{#1}}}
    \newcommand*{\ang}[0]{\;\text{\AA}}
    \newcommand*{\mum}[0]{\;\upmu \mathrm{m}}
    \newcommand*{\at}[1]{\left.#1\right|}

@@

    \DeclarePairedDelimiter\bra{\langle}{\rvert}
    \DeclarePairedDelimiter\ket{\lvert}{\rangle}
    \DeclarePairedDelimiter\abs{\lvert}{\rvert}
    \DeclarePairedDelimiter\ev{\langle}{\rangle}
    \DeclarePairedDelimiter\p{\lparen}{\rparen}
    \DeclarePairedDelimiter\s{\lbrack}{\rbrack}
    \DeclarePairedDelimiter\z{\lbrace}{\rbrace}

@@

Another formula given by Sutherland 2011 is $\pd{\ev*{E}}{t} = v_{g, z}
\ev*{E}$ where we average over $x$ one wavelength. I wasn't able to show
that these two agree in general; Sutherland's formula seems to be derived for a
traveling wavepacket whereas we have no such thing. We do not use this
expression since the Landau \& Lifshitz expression works well.

@@

One can imagine, for this matter, that $\pd{}{t}$ can be wrapped in the $L_{ij}$
operator and the timestepping computed implicitly; I believe this is what
Dedalus does, implicitly timestepping the left hand side and explicitly the
inhomogeneous terms.

Both of these procedures require $\bm{L}$ linear operator or $\pd{}{t} + \bm{L}$
to be full rank, or invertible. This is enforced analytically by requiring the
same number of BCs as derivatives, and numerically this is little different.
This is why it is possible to have components of $Q$ that are not differentiated
$\partial_t$ but still have well-defined time evolutions; the only condition
required is that the linear terms of the PDE make up a full-rank operator.
Hence, in our linear incompressible equations, we permitted use of $\vec{\nabla}
\cdot \vec{u}_1 = 0$ in lieu of a $\pd{P}{t}$ explicit equation of motion; so
long as the gauge for $P$ (which only appears in derivatives) is properly set,
the operator is full rank and numerically well-defined.

@@

\section{Damping Zone}

Since our code is a spectral code, instead of using the quadratic damping zone
presented from before (whose second derivative is discontinuous over the entire
domain), we prefer instead an $\arctan$ profile, such that
\begin{equation}
    \Gamma(z) \propto 0.5 * \p*{
        2 + \tanh C\p*{\frac{z - z_t}{L_z - z_t}}
            + \tanh C\p*{\frac{z - z_b}{z_b}}}.
\end{equation}
This is a profile that mostly vanishes $z \in [z_b, z_t]$ but is unity for $z
\lesssim z_b, z \gtrsim z_t$, with steepness of transition governed by factor
$C$. While this profile may have reflection issues if $C$ is chosen to be too
steep, it is smooth and so has exponentially convergent spectral expansion and
empirically seems to do alright.

It is moreover prudent to note that, since implicit calculations are expensive,
as long as the damping timescale $\Gamma^{-1}$ is everywhere resolved it is
cheaper to put $\Gamma \rho_1, \Gamma \vec{u}_1$ on the right hand sides of
their equations in Dedalus, which is timestepped explicitly.

@@

Let us derive the resultant equations of motion. In the interest of maintaining
a divergence-free velocity field, we place the driving term on the density
equation. Since the forcing is purely to excite waves and not physical (for the
time being), it should not worry us that driving the density equation is
ill-justifiable physically.

Consider the linearized system of equations (to examine the excited perturbation
in the linear legime)
\begin{subequations}\label{se:vol_drive}
    \begin{align}
        \vec{\nabla} \cdot \vec{u} &= 0,\\
        \pd{\vec{u}_1}{t} + \frac{\vec{\nabla}P}{\rho_0}
            + \frac{\rho_1 \vec{g}}{\rho_0}
            &= 0,\\
        \pd{\rho_1}{t} - \frac{u_{1z} \rho_0}{H}
            &= F(x, z, t),\\
        F(x, z, t) &= \Re Fe^{-\frac{(z - z_0)^2}{2\sigma^2}}
            e^{i\p*{k_xx - \omega t}}.
    \end{align}
\end{subequations}
For the linearized system, we can immediately substitute $\pd{}{t} \to -i\omega,
\pd{}{x} \to ik_x$, giving
\begin{align*}
    \pd{u_{1z}}{z} + ik_xu_x &= 0,\\
    -i\omega u_x + \frac{ik_xP}{\rho_0} &= 0,\\
    -i\omega u_{1z} + \frac{1}{\rho_0}\pd{P}{z} + \frac{\rho_1 g}{\rho_0} &= 0,\\
    -i\omega \rho_1 - \frac{u_{1z}\rho_0}{H} &= F.
\end{align*}
This can be recast solely in terms of $u_{1z}$ as
\begin{align*}
    -i\omega u_{1z} + \frac{\omega}{k_x}\p*{-\frac{1}{ik_x}\ptd{u_{1z}}{z}
        + \frac{1}{ik_xH}\pd{u_{1z}}{z}} - \frac{u_{1z}g}{Hi\omega}
        - \frac{Fg}{i\omega \rho_0} = 0,\\
    \omega^2\p*{
        k_x^2u_{1z} + \frac{1}{H}\pd{u_{1z}}{z} - \ptd{u_{1z}}{z}
    } - u_{1z}N^2k_x^2 - \frac{Fgk_x^2}{\rho_0} = 0.
\end{align*}
Note that if $\pd{}{z} \to 1/2H, F = 0$ then this simplifies to the expected
dispersion relation $\omega^2 = \frac{N^2k_x^2}{k_x^2 + k_z^2 +
\frac{1}{4H^2}}$.

If we approximate $F(x, z, t) \approx F_0\delta\p*{z - z_0} e^{i(k_xx - \omega t)}$
for the time being, then clearly the only term that can be equally singluar is
the $\ptd{}{z}$ term, leaving us with
\begin{equation}
    -\omega^2 \ptd{u_{1z}}{z} - \frac{Fgk_x^2}{\rho_0} = 0.
\end{equation}
Since $u_{1z}$ both above and below $z_0$ satisfy the homogeneous PDE, which has
plane wave solutions with an $e^{z/2H}$ envelope where $k_zH \gg 1$, we can
integrate about a small domain around $z_0$ and obtain
\begin{equation}
    \at{\pd{u_{1z}}{z}}_{z > z_0} - \at{\pd{u_{1z}}{z}}_{z < z_0}
        = -F_0\frac{gk_x^2}{\rho_0\omega^2}e^{i(k_xx - \omega t)}.
\end{equation}
It should then be apparent that the $u_{1z}$ solution is
\begin{equation}
    u_{1z} = \frac{F_0gk_x^2}{\rho_0(z = z_0)\omega^2}\frac{1}{2ik_z} \times
    \begin{cases}
        -e^{(z - z_0)/2H}e^{i(k_xx + k_z(z - z_0) - \omega t)} & z > z_0\\
        e^{(z - z_0)/2H}e^{i(k_xx + k_z(z - z_0) - \omega t)} & z < z_0\\
    \end{cases}.
\end{equation}

Given that we are actually using a Gaussian $F(x, z, t) \propto F_0e^{-\frac{(z
- z_0)^2}{2\sigma^2}}$, the integral under $F(z)$ is actually
$F_0\sqrt{2\pi}\sigma$ rather than just $F_0$. Finally, since a delta function
is flat in spatial frequency space while a Gaussian tapers off with width
$1/\sigma$ ($\sigma$ is the spatial width, so $1/\sigma$ is the spatial
frequency width), we should expect an overall multiplicative factor of
$e^{-k_z^2\sigma^2 / 2}$ in the $k_z$ spatial component of $F(x, z, t)$ compared
to the delta function case. Thus, we finally obtain
\begin{equation}
    \abs*{u_{1z}(z = z_0)} = \frac{F_0\sqrt{2\pi}\sigma gk_x^2}{
        2\rho_0\p*{z = z_0}\omega^2 k_z}e^{-k_z^2\sigma^2/2}.
\end{equation}

@@

Note: There exist papers on the Courant stability condition of Chebyshev
spectral methods which cite exactly the $N^{-2}$ minimum timestep scaling we can
infer from the above. But if our field is small near the boundaries and is only
interesting in a region isolated from the boundary (e.g.\ with our damping
zones) then the bound used in the papers scales more like $N^{-1}$, which is
much more desirable.
